### 2016-03-21 Manuscript/SW_early_life_telomeres.Rmd
More analyses

**Strong cohort effects on early-life telomere length in a wild population**

@@

###Abstract
Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Thus, it is possible that telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival. We tested how telomere dynamics are related to spatio-temporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(juv)` juveniles from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years. Despite cross-sectional data showing that telomere length decreases with age, and most that telomere length decreases most rapidly early in life, using longitudinal data we found that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero. Juvenile telomere length varied markedly among cohorts, and this cohort variation was related to the season (birds born in winter seasons had longer telomeres), and to population density (birds bird in years with higher populaiton density had longer telomered). Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening. Finally, we found no relationship between early-life and survival to adulthood. Our data suggest that telomeres are subject to strong population-level variation, but suggest that the resolution of telomeres as a useful biomarker of individual-level cost is likely to be age-dependent, and in many cases limited.

@@

All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment. A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples. However, this did not affect final telomere length values (see below). We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate. Averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material), excluding samples with Cq values differing by >0.5. We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001]. We chose to use RTL rather than continuing with the previously used method for calculating absolute telomere length [@Barrett2012], as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable. Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.

@@

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subsject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models, with individual ID as a random effect. For the analysis of $\Delta$age we we only included birds who had been sampled at least twice. We then explored how telomere length varied within the first year of life (telomere length vs age in months) and over time (telomere length vs cohort ID). For this analysis we had bvery few birds with multiple samples, so we used linear models including only the first sample for each individual.

@@

We used general linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).

@@

#####Early-life RTL
Considering the entire Seychelles warbler lifespan, RTL decreased with age (t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`; Fig. 1A), and with $\Delta$age (t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig. 1B). RTL also decreased with age within the first year of life (i.e. excluding adult samples; R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; t = `r round2(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,3])`; P `r round3(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,4])`). Early-life RTL varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`), and cohort-level variation in juvenile RTL (taken as the median for each cohort) was positively related to adult population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Density,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 1C). Variation in cohort-level RTL was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~TQ,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), or island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Insect,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`).

The top model explaining variation in early-life RTL contained age, sex, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1). The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig. 2A). RTL was higher in winter compared to summer seasons (Fig. 2B). Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig. 2C). Males had slightly longer telomeres than females, prarticularly at the fledgling and subadult stages (Fig. 2D), although this effect was weak.

#####Early-life telomere shortening
Longitudinal data showed that early-life RTL was positively, but weakly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig. 3A). $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`). In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase. In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`). However, for fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`). In flesdglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B). Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life. We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings.

$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts was low (Fig. 3C). Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2), and confidence limits for all explanatory terms overlapped zero (Fig. 4A).

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+Tarsus+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).

@@

###Discussion
Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere length and survival in a closed population of Seychelles warblers. We found that, as in other studies, telomeres are lost at the greatest rate very early in life. We found that telomere dynamics are subject to strong seasonal and cohort effects, but that temporal variation in RTL is not related to any of the environmental variables tested. Within seasons, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.

Our study adds the the growing number showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012]. Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig. 2) most likely an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976]. Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected. However, despite our clear finding that population-level RTL decreases with age throughout the warbler lifespan (Fig. 1), we found that individual-level telomere shortening can only be detected  very early in life (Fig. 3). This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nestbox [@Boonekamp2014] conditions, in which individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood. However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found a weak correlation between juvenilt and adult RTL, and a highly complex pattern of telomere dynamics across the entire lifespan, with periods of early-life telomere shortening followed by selective disappearance and apparent telomere lengthening as individuals age [@Fairlie2015]. Indeed, the complex relationship between RTL and age in our study is strikingly similar to that reported by Fairlie *et al.* [-@Fairlie2015]. It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie 2015]. As early-life telomere dynamics are investigated in more systems, the causes of this variation should become clear.

Perhaps the clearest result from our study is that RTL varies among cohorts. Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, these have been limited in the number of seasons [@Mizutani2013; @Watson2015; but see @Fairlie2015]. The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics does occur at the population level, over decadal time periods, and that this variation was related to variation in the environment (Fig. 1C). The difference in RTL between birds born in summer and winter seasons is more surprising, especially given the strength and directionality of the effect. We find that birds born in winter seasons having longer telomeres, despite the . One possible explanation for this is that parents in good condition are more likely to breed

We suspect that the novelty in our finding temporal variation in telomere dynamics within a population is more due to a lack of available long-term datasets with telomere screening, rather than the Seychelles warbler being unique. Indeed, the population on Cousin is relatively stable in comparison with other populations, which can undergo large fluctuations in size [e.g. @Coulson2001]. If our findings are replicated in other systems and population-level variation in early-life telomere dynamics is common in nature, this has potential consequences for our understanding of telomere dynamics in natural populations. In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process. More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.

>>>>

**Early-life conditions, telomere dynamics and survival: insights from a long-term island bird study**

@@

###Abstract
1. Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Thus, it is possible that telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival.

2. We tested how telomere dynamics are related to spatio-temporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(juv)` juveniles from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years.

3. We found that at the population level, telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life. However, at the individual level we found only a very weak relationship between juvenile and adult telomere length, and that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero.

4. Juvenile telomere length varied markedly among cohorts, and birds born in years with higher populaiton density had longer telomeres. Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening. We also found no relationship between early-life telomere dynamics and survival later in life.

5. Our data suggest that telomeres are subject to strong population-level variation, but suggest that the resolution of telomeres as a useful biomarker of individual-level cost is likely to be age-dependent, and in many cases limited.

@@

All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with two amendments. A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples. However, this did not affect final telomere length values (see below). Second, to avoid bacth effects, we randomised samples among plates. We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate. Averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material), excluding samples with Cq values differing by >0.5. We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001]. We chose to use RTL rather than continuing with the previously used method for calculating absolute telomere length [@Barrett2012], as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable. Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.

@@

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subsject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models, with individual ID as a random effect. For the analysis of $\Delta$age we we only included birds who had been sampled at least twice. We then explored how telomere length varied within the first year of life (telomere length vs age in months) and among cohorts (telomere length vs cohort ID), again using linear mixed models with individual ID as a random effect.

@@

We used general linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (major or minor) and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).

@@

#####Early-life RTL
Considering the entire Seychelles warbler lifespan, RTL decreased with age (LMM: t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`), although there were clear increases in cross-sectional telomere length after one year of age and at several points later in life (Fig. 1A). RTL also decreased with age within individuals (i.e. with $\Delta$age; t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig. 1B). RTL also decreased with age within the first year of life (i.e. excluding adult samples; t = `r round2(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=juv,random=~1|BirdID))$tTable[2,5])`).

Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`). Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C). Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`). RTL did not differ in birds born in minor compared to major seasons (t = `r round2(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`; Fig. 2A, 2C).

Within cohorts, the top model explaining variation in early-life RTL contained age, sex, tarsus length, and the interaction between age and tarsus length (Table S1). The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Tarsus+Agemonths*Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig. 3A). Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig. 3B). Males had slightly longer telomeres than females, particularly at the fledgling and subadult stages, although this effect was weak  (Fig. 3C).

#####Early-life telomere shortening
Longitudinal data showed that early-life RTL was positively, but weakly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig. 4A). $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`). In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase. In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`). However, in fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`; Fig. 4B, Fig. S1). In both fledglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B). Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life. We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings (N = `r nrow(subset(Loss,Agemonths == 1))`.

$\Delta$RTL did not differ significantly among cohorts of nestlings (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low. Model selection revealed that the top model explaining $\Delta$RTL in nestlings contained sex and tarsus length (Table S2). This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE))-AICc(lmer(Loss~Sex+Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE)))`), although confidence limits for both the sex and tarsus effects overlapped zero (Fig. 5A). $\Delta$RTL was negatively related to tarsus length, and higher in females, indicating higher rates of telomere loss in smaller nestlings, and in males.

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).

@@

###Discussion
Here we use one of the largest datasets to date to assess the relationships between early-life conditions, telomere length and survival in a closed population of Seychelles warblers. We found that, as in other studies, telomeres are lost at the greatest rate  early in life. We found that telomere dynamics are subject to strong cohort effects, and that temporal variation in RTL is population density and season. Within cohorts, we found no evidence that the ecological or social environment influenced telomere dynamics, and we found no evidence that early-life telomere length is related to juvenile or later-life survival.

Our study adds the the now substantial body of literature showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2015]. Our finding that tarsus length was related to juvenile RTL in an age-dependent manner (Fig. 3) most likely represents an additional age effect - in passerine birds tarsus length is correlated with age during the nestling stage, but not after fledging [@Ricklefs1976]. Telomere loss is most rapid early in life due to ongoing growth [@Frenck1998], and a negative correlation between RTL and body size in early life is therefore expected [see also @Ringsby2015]. However, despite our clear finding that population-level RTL decreases with age throughout the warbler lifespan (Fig. 1), we found that at the individual-level, juvenile and adult telomere length were weakly correlated, and that telomere shortening can only be detected very early in life (Fig. 4). This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nestbox [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood. However, in accordance with our study, a recent study on Soay sheep (*Ovis aries*) found a weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the entire lifespan [@Fairlie2015]. Indeed, the complex relationship between RTL and age in our study, with periods of early-life telomere shortening followed by selective disappearance and/or apparent telomere lengthening as individuals age is strikingly similar to that reported by Fairlie *et al.* [-@Fairlie2015]. It is too soon to speculate as to what may be generating this variation among species, but one possibility is that environmental variability itself affects the resolution with which we can observe telomere shortening in natural populations [see also @Fairlie 2015]. As early-life telomere dynamics are investigated in more systems, the causes of this variation should become clear.

Perhaps the clearest result from our study is that RTL varies among cohorts. Very few studies have shown that temporal variation in telomere dynamics occurs in natural populations, and these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2015]. The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods. Moroevoer, becuase of the larger number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in the environment (Fig. 1C). We may expect a negative relationship between population density and RTL, as increased competition for resources is known to effect juvenile telomere length. However, We suggest that population density acts as an indicator of overall island 'quality' for Seychelles warblers. This can clearly be seen from data from the 1980s and 1990s, when adult population size on Cousin increased as habitat quality improved [@Komdeur2005]. Population size has fluctuated around 300 since the late 1990s (Fig. 2B), and captures this more accurately than measures of insect abundance or territory quality.

Within cohorts, we found no evidence that any of the social or environmental variables we measured affect juvenile RTL. This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival. Possible explanations for this finding include i) temporal variation in RTL in our data masks our ability to detect spatial trends, ii) within-cohort variation in RTL is explained by a variable that we have not included in our analyses, or iii) that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects. The first of these explanations seems the least likely, as within-cohort variation in RTL greatly exceeds among-cohort variation (Fig. 1C).
